Seasonality in dominant upper-ocean dynamics at submesoscales

### 2016-05-09 writeup/rocha_etal.tex
Adds details of numerical method
Adds more content to secs 3 and 4

A key aspect of the simulation analyzed here is that it is forced by
the 16 most significant tidal components.
Barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep{rocha_etal2016}. Hence, tidal forcing fundamentally differentiates our analysis
from other modeling studies that investigated upper dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}.

The region of focus is the northwest Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have
shown strong seasonality in the quasi-gesotrophic surface velocity field.
We analyze a sub-domain of the LLC2160 simulation of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N. We focus on the surface velocity field
and near-surface stratification.

@@

We calculate the vertical vorticity, horizontal divergence, and the magnitude
of the lateral rate of strain:
\begin{equation}\label{vort}
\zeta \equiv v_x - u_y\, ,
\end{equation}
\begin{equation}\label{div}
\delta \equiv u_x + v_y\, ,
\end{equation}
and
\begin{equation}\label{strain}
S \equiv [(v_x+u_y)^2 + (u_x-v_y)^2]^{1/2}\, ,
\end{equation}

@@

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.9\textwidth]{figs/fig2.pdf}
  \caption{Time series of horizontal velocity diagnostics at the surface
          (a through d) and stratification in the upper 50 m (e).}
  \label{fig2}
  \end{center}
\end{figure*}

@@

@@

\subsection{The high-frequency signal}
Here we argue that those divergent motions are likely inertia-gravity waves.
We can show a couple of spectra and argue that they roughly satisfy polarization
relations.

We can also try and show some observations here. Any cruises with ADCP data.
While it may be hard to have enough data for the errorbars to be small, a
rough consistency may be better than nothing.

Perhaps a mooring data should show strong seasonal modulation at high-frequencies?

We must be able to present some observational evidence that the model is not
misleading at high-frequencies.

\subsection{Geostrophic velocities from SSH}
Here we argue that high-frequency (supra-daily) flows significantly project
on the sea-surface, and thus estimation of submesoscale (10-100 km) geostrophic
velocity from sea-surface height (SSH) is not warranted. Calculating spectral
KE fluxes from both velocity and geostrophic velocity (from SSH) would
contrast with the results in Sasaki et al.

>>>>

A key aspect of the simulation analyzed here is that it is forced by
the 16 most significant tidal components.
Barotropic tides interact with topography and generate internal
tides that project on mesoscales to submesoscales
\citep{rocha_etal2016}. Hence, tidal forcing fundamentally differentiates our analysis
from recent modeling studies that investigated upper dynamics
and its seasonality \citep{sasaki_etal2014,qiu_etal2014}.

The region of focus is the northwest Pacific, in the vicinities of the Kuroshio
Extension, where previous studies have
shown strong seasonality in the quasi-gesotrophic surface velocity field.
We analyze a sub-domain of the LLC2160 simulation of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (figure \ref{fig1}a-b).

@@

Focusing on the lateral gradients of the horizontal velocity vector.
The lateral gradients are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$,
horizontal divergence $\delta \equiv u_x + v_y$, and the magnitude
of the lateral rate of strain $S \equiv [(v_x+u_y)^2 + (u_x-v_y)^2]^{1/2}$.
These quantities highlight the submesoscale structures in the flow
and hint on their dynamics.

Figures \ref{fig1}a-b show snapshots of vertical vorticity $\zeta$ in early Spring
(April 15) and Fall (October 15). Clearly, a strong seasonality in vorticity
exists: large values of fine-grained vertical vorticity are
observed in early Spring with maximum values as high as $0.9f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about 0.25. In early
Fall, the situation is the opposite: the vertical vorticity is coarse-grained,
and its local maximum and RMS is no larger than $0.3f$ and $0.1f$, respectively.
Indeed, the seasonal cycle in vorticity/strain is very strong (Figure~\ref{fig2}):
the RMS vertical vorticity is about twice as large in late Winter/early Spring
than it is in late Summer/early Fall. This picture is consistent with
recent studies: shallow baroclinic instabilities energize the submesoscales
in late Winter, drawing from the available potential energy stored in deep mixed
layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

@@

The bulk of the vertical vorticity is associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields reduces the vertical
vorticity by less than 20$\%$ and the seasonal cycle remains strong (see
green lines in figure \ref{fig2}). Most of this seasonal cycle is associated
with submesoscale flows: low-pass filtering the velocity field
with a cut-off scale of 100 km reduces the RMS vorticity by at least 40$\%$,
yielding bulk statistics of velocity gradient roughly consistent with AVISO
geostrophic velocity data (see lines blue and black in figure \ref{fig2}).

The horizontal divergent part of the flow is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows project are supra-inertial and project
on scales smaller than 100 km: daily-averaging the velocity fields or low-pass
filtering (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Furthermore, the seasonal cycle in horizontal divergence is out-of-phase with that
of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.

 \begin{figure*}[ht]
   \begin{center}
     \includegraphics[width=.9\textwidth]{figs/fig2.pdf}
  \caption{Time series of the root-mean-square of kinetic energy (a), vorticity
  (b), lateral strain (c), horizontal divergence (d), and stratification in the upper 50 m (e).}
  \label{fig2}
  \end{center}
\end{figure*}

\section{Projection on horizontal scales}

% Here we argue that those divergent motions are likely inertia-gravity waves.
% We can show a couple of spectra and argue that they roughly satisfy polarization
% relations.
%
% We can also try and show some observations here. Any cruises with ADCP data.
% While it may be hard to have enough data for the errorbars to be small, a
% rough consistency may be better than nothing.
%
% Perhaps a mooring data should show strong seasonal modulation at high-frequencies?
%
% We must be able to present some observational evidence that the model is not
% misleading at high-frequencies.

% Here we argue that high-frequency (supra-daily) flows significantly project
% on the sea-surface, and thus estimation of submesoscale (10-100 km) geostrophic
% velocity from sea-surface height (SSH) is not warranted. Calculating spectral
% KE fluxes from both velocity and geostrophic velocity (from SSH) would
% contrast with the results in Sasaki et al.

@@

\section{Projection on time scales}
To better understand the dynamics the projection of these flows on horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance.

@@
